Long-form recording of infant body position in the home using wearable inertial sensors

### 2023-03-16 manuscript.Rmd
more discussion

Across three analyses, we demonstrated the validity of long-form recordings of infant body position using wearable inertial sensors. Both group and individual models trained during the proximal comparison period performed well on testing data collected from the same recording period. Most important, accuracy was consistently strong later in the recording when behavior was completely unsupervised: Human-coded and model-predicted durations of body positions during the distal comparison period were highly correlated, even when narrowing to the scale of 10 minutes. Examining full recordings showed that the new method allows us to capture more data (median = `r apa_num(median(coverage$hours_recorded))` hours of awake behavior) than is typical with video methods (1-2 hours) while simultaneously reducing the human labor needed to annotate it. Ultimately, age differences in body position mirrored past findings that employed other methods, suggesting that the end product is suitable for describing developmental changes in motor behavior.

@@

Counterintuitively, the two outlier participants increased our confidence that the body position classification method works as intended. At first glance, seeing low agreement rates for two sessions would suggest that the models perform poorly at predicting some *infants'* behavior. Instead, we found that for those infants, confusion was restricted to two positions during a single, long-lasting event and the other three body positions continued to be classified correctly. In other words, the models failed to predict a particular *event* for those two infants. Unfortunately, those events lasted a long time. In the case of confusion between supine and sitting for the infant who was reclined, the confusion likely stems from the difficult of categorizing behaviors that do not always fit neatly into discrete classes. Even to a human observer, there are often "edge" cases when scoring infant behavior, which is why inter-rater reliability is never perfect.

## Group and Individual Models

No clear winner. Higher raw accuracy/kappa for individual models; some (but not all) better by-class performance for Kappa's. But group models better for prone; more sensitive for prone and for held; barring outliers; group has some better part 2 correlations depending on the class. Notably, worst outlier in aggregate age difference data was in an individual model.

Big advantages of group model are that 1) trained on more data (so likely a bit more robust), 2) less dependent on getting the perfect training set, 3) between individual differences are due to behavior, not due to choice of model, 4) can apply to any infant, 5) can apply to an infant for whom we have no video. Individual models might capture some idiosyncracies of movement (or device use) that are hard to get from big model that homogenizes behavior. Individual models are a nice choice for rapid prototyping/piloting/feasibility. Can get proof of concept on a single participant or a new kind of code (loc; stroller use; etc.). Clinical cases too.

Tell some things we did but didn't report. Tried modeling each age separately but there was no overall improvement in performance. Categories are so broad and include less/more mature forms of behavior. Not independent sitting, but all kinds of sitting. Tried other ML algorithms. Tried fewer sensors.

>>>>

Across three analyses, we demonstrated the validity of long-form recordings of infant body position using wearable inertial sensors. Both group and individual models trained during the proximal comparison period performed well on testing data collected from the same recording period. Most important, accuracy was consistently strong later in the recording when behavior was completely unsupervised: Human-coded and model-predicted durations of body positions during the distal comparison period were highly correlated, even when narrowing to the scale of 10 minutes. Examining full recordings showed that the new method allows us to capture more data (median = `r apa_num(median(coverage$hours_recorded))` hours of awake behavior) than is typical with video methods (1-2 hours) while simultaneously reducing the human labor needed to annotate it. Ultimately, age differences in body position mirrored past findings that employed other methods, suggesting that the outcome of full-day body position recordings is suitable for describing developmental changes in motor behavior.

@@

Counterintuitively, the two outlier participants increased our confidence that the body position classification method works as intended. At first glance, seeing low agreement rates for two sessions would suggest that the models perform poorly at predicting some *infants'* behavior. Instead, we found that for those infants, errors were restricted to two positions during a single, long-lasting event, while the other three body positions continued to be classified correctly. In other words, the models failed to predict a particular *event* for each of the two infants. Unfortunately, those events lasted a long time leading to poor overall agreement. In the case of confusion between supine and sitting for the infant who was reclined, the confusion likely stems from the difficult of categorizing behaviors that do not always fit neatly into discrete classes (e.g., continuous body position angles categorized as either supine or sitting).

## Benefits and Drawbacks of Different Modeling Approaches

Throughout the paper, we compared results from two modeling approaches: group models that included all participants' data in the training set versus individual models that used only one participant's data. When considering different metrics, testing periods, body positions, as well as the logistical benefits of each, there is no clear winner. Below, we discuss the pros and cons to each approach so that researchers might decide what works best for their intended application.

Overall accuracy and kappa values were better in individual models compared with group models when collapsed across body positions in the proximal period. However, within-class performance did not always favor individual models---prone predictions were better for group models, and held predictions were almost identical. In the distal comparison (after removing outliers), overall correlations were nearly identical. Within classes, group models had an advantage for prone, upright, and held, whereas individual models had an advantage for sitting and supine. Possibly, individual models are better suited for capturing unique aspects of the devices used for sitting and reclining that lead to better performance in those classes. Finally, age effects (either by group or by age correlation) were nearly identical across the two methods, suggesting that either method would lead to the correct conclusion regarding developmental changes in body position. Outliers were present in both models' predictions: group models produced three sessions with the worst overall accuracy, but individual models produced the single worst kappa. The outliers for the distal comparison period appeared in both models' predictions. For aggregated full-day body position, the most blatant outlier (a young infant held > 60% of the day) resulted from an individual model. The choice of model might depend on which behavior is most important to capture for a given research question.

Aside from differences in validity, researchers may favor one modeling approach based on logistical concerns. Group models have several major advantages. First, by definition they are trained on more data, which might make predictions more consistent. Second, group models remove the need to get optimal training data from each participant. It can be difficult to elicit every behavior of interest in each infant; individual models require that every behavior that will be later predicted was displayed by each individual infant. In fact, group models remove the need to get *any* training data for a particular infant, meaning that they can be applied to any infant regardless of whether they have training data. Finally, applying a group model means that individual differences between infants are due to their behavior being classified in a consistent way, rather than having a separate set of rules for predicting each individual. However, given the variety in infant behavior, individual models are more agile in that they can capture the unique and unexpected. Likewise, individual models can be an excellent choice for rapid prototyping and pilot testing. A researcher can get proof-of-concept data from a single participant or a new kind of classification scheme (i.e., locomotion, restrain in device) without annotating an entire sample. Individual models may also be applied in clinical cases where infants vary widely in their motor abilities and/or use of assistive devices.

Beyond group and individual models, there were many degrees of freedom in our choice of how best to model body position. For the sake of brevity, we did not report every possible variation, only the best approach. However, our openly shared data mean that researchers can experiment with different sets of motion features, different sets of training data, different machine learning algorithms, and different sets of hyperparameters. As in past work [@AiraksinenRasanen2020], performance is best when using all four sensors and degrades when using any subset of two (i.e., hips only, left leg only). We tested but did not report models that used either only accelerometer data and only gyroscope data; in both cases, performance is degraded compared with models that use both types of motion. We also tried different machine learning algorithms (XGBoost, Tabnet), but found that random forest models performed the best. Finally, we tried modeling each age group separately, but found no boost in performance by tailoring to the particular age group.
<!-- DATA CITE -->

### 2023-03-16 manuscript.tex
more discussion

Across three analyses, we demonstrated the validity of long-form recordings of infant body position using wearable inertial sensors. Both group and individual models trained during the proximal comparison period performed well on testing data collected from the same recording period. Most important, accuracy was consistently strong later in the recording when behavior was completely unsupervised: Human-coded and model-predicted durations of body positions during the distal comparison period were highly correlated, even when narrowing to the scale of 10 minutes. Examining full recordings showed that the new method allows us to capture more data (median = 6.15 hours of awake behavior) than is typical with video methods (1-2 hours) while simultaneously reducing the human labor needed to annotate it. Ultimately, age differences in body position mirrored past findings that employed other methods, suggesting that the end product is suitable for describing developmental changes in motor behavior.

@@

Counterintuitively, the two outlier participants increased our confidence that the body position classification method works as intended. At first glance, seeing low agreement rates for two sessions would suggest that the models perform poorly at predicting some \emph{infants'} behavior. Instead, we found that for those infants, confusion was restricted to two positions during a single, long-lasting event and the other three body positions continued to be classified correctly. In other words, the models failed to predict a particular \emph{event} for those two infants. Unfortunately, those events lasted a long time. In the case of confusion between supine and sitting for the infant who was reclined, the confusion likely stems from the difficult of categorizing behaviors that do not always fit neatly into discrete classes. Even to a human observer, there are often ``edge'' cases when scoring infant behavior, which is why inter-rater reliability is never perfect.

\hypertarget{group-and-individual-models}{%
\subsection{Group and Individual Models}\label{group-and-individual-models}}

No clear winner. Higher raw accuracy/kappa for individual models; some (but not all) better by-class performance for Kappa's. But group models better for prone; more sensitive for prone and for held; barring outliers; group has some better part 2 correlations depending on the class. Notably, worst outlier in aggregate age difference data was in an individual model.

Big advantages of group model are that 1) trained on more data (so likely a bit more robust), 2) less dependent on getting the perfect training set, 3) between individual differences are due to behavior, not due to choice of model, 4) can apply to any infant, 5) can apply to an infant for whom we have no video. Individual models might capture some idiosyncracies of movement (or device use) that are hard to get from big model that homogenizes behavior. Individual models are a nice choice for rapid prototyping/piloting/feasibility. Can get proof of concept on a single participant or a new kind of code (loc; stroller use; etc.). Clinical cases too.

Tell some things we did but didn't report. Tried modeling each age separately but there was no overall improvement in performance. Categories are so broad and include less/more mature forms of behavior. Not independent sitting, but all kinds of sitting. Tried other ML algorithms. Tried fewer sensors.

>>>>

Across three analyses, we demonstrated the validity of long-form recordings of infant body position using wearable inertial sensors. Both group and individual models trained during the proximal comparison period performed well on testing data collected from the same recording period. Most important, accuracy was consistently strong later in the recording when behavior was completely unsupervised: Human-coded and model-predicted durations of body positions during the distal comparison period were highly correlated, even when narrowing to the scale of 10 minutes. Examining full recordings showed that the new method allows us to capture more data (median = 6.15 hours of awake behavior) than is typical with video methods (1-2 hours) while simultaneously reducing the human labor needed to annotate it. Ultimately, age differences in body position mirrored past findings that employed other methods, suggesting that the outcome of full-day body position recordings is suitable for describing developmental changes in motor behavior.

@@

Counterintuitively, the two outlier participants increased our confidence that the body position classification method works as intended. At first glance, seeing low agreement rates for two sessions would suggest that the models perform poorly at predicting some \emph{infants'} behavior. Instead, we found that for those infants, errors were restricted to two positions during a single, long-lasting event, while the other three body positions continued to be classified correctly. In other words, the models failed to predict a particular \emph{event} for each of the two infants. Unfortunately, those events lasted a long time leading to poor overall agreement. In the case of confusion between supine and sitting for the infant who was reclined, the confusion likely stems from the difficult of categorizing behaviors that do not always fit neatly into discrete classes (e.g., continuous body position angles categorized as either supine or sitting).

\hypertarget{benefits-and-drawbacks-of-different-modeling-approaches}{%
\subsection{Benefits and Drawbacks of Different Modeling Approaches}\label{benefits-and-drawbacks-of-different-modeling-approaches}}

Throughout the paper, we compared results from two modeling approaches: group models that included all participants' data in the training set versus individual models that used only one participant's data. When considering different metrics, testing periods, body positions, as well as the logistical benefits of each, there is no clear winner. Below, we discuss the pros and cons to each approach so that researchers might decide what works best for their intended application.

Overall accuracy and kappa values were better in individual models compared with group models when collapsed across body positions in the proximal period. However, within-class performance did not always favor individual models---prone predictions were better for group models, and held predictions were almost identical. In the distal comparison (after removing outliers), overall correlations were nearly identical. Within classes, group models had an advantage for prone, upright, and held, whereas individual models had an advantage for sitting and supine. Possibly, individual models are better suited for capturing unique aspects of the devices used for sitting and reclining that lead to better performance in those classes. Finally, age effects (either by group or by age correlation) were nearly identical across the two methods, suggesting that either method would lead to the correct conclusion regarding developmental changes in body position. Outliers were present in both models' predictions: group models produced three sessions with the worst overall accuracy, but individual models produced the single worst kappa. The outliers for the distal comparison period appeared in both models' predictions. For aggregated full-day body position, the most blatant outlier (a young infant held \textgreater{} 60\% of the day) resulted from an individual model. The choice of model might depend on which behavior is most important to capture for a given research question.

Aside from differences in validity, researchers may favor one modeling approach based on logistical concerns. Group models have several major advantages. First, by definition they are trained on more data, which might make predictions more consistent. Second, group models remove the need to get optimal training data from each participant. It can be difficult to elicit every behavior of interest in each infant; individual models require that every behavior that will be later predicted was displayed by each individual infant. In fact, group models remove the need to get \emph{any} training data for a particular infant, meaning that they can be applied to any infant regardless of whether they have training data. Finally, applying a group model means that individual differences between infants are due to their behavior being classified in a consistent way, rather than having a separate set of rules for predicting each individual. However, given the variety in infant behavior, individual models are more agile in that they can capture the unique and unexpected. Likewise, individual models can be an excellent choice for rapid prototyping and pilot testing. A researcher can get proof-of-concept data from a single participant or a new kind of classification scheme (i.e., locomotion, restrain in device) without annotating an entire sample. Individual models may also be applied in clinical cases where infants vary widely in their motor abilities and/or use of assistive devices.

Beyond group and individual models, there were many degrees of freedom in our choice of how best to model body position. For the sake of brevity, we did not report every possible variation, only the best approach. However, our openly shared data mean that researchers can experiment with different sets of motion features, different sets of training data, different machine learning algorithms, and different sets of hyperparameters. As in past work (Airaksinen et al., 2020), performance is best when using all four sensors and degrades when using any subset of two (i.e., hips only, left leg only). We tested but did not report models that used either only accelerometer data and only gyroscope data; in both cases, performance is degraded compared with models that use both types of motion. We also tried different machine learning algorithms (XGBoost, Tabnet), but found that random forest models performed the best. Finally, we tried modeling each age group separately, but found no boost in performance by tailoring to the particular age group.
